Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-03 sections/related.tex
updates
Update on Overleaf.

\paragraph{ Compilers for Structured Data:}
Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
%
The TACO compiler originally supported just plain einsum computations, but has been extended several times to support (single dimensional) local tensors, breaks via semi-rings, windowing, tiling, and convolution~\cite{kjolstad_tensor_2017, kjolstad_tensor_2019, senanayake2020sparse, henry_compilation_2021,won2023unified}.
%
Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures and tree like structures~\cite{kjolstad_tensor_2017, chou2018format, chou2022compilation}.
%
The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}.
%
Similarly, SDQL originally supported only hash tables, but has been extended with one format language that allows one to specify tensors as queries on several base storage types~\cite{schleich2023optimizing} and via another system that optimizing programs by taking advantage of predicates describing symmetries and other structures on a tensor~\cite{ghorbani2023compiling}.
%
Taichi is another language with a separate program and data structure specification, but Taichi focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
%
Outside of these works, the window of comparison fogs up.
%
The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
%
Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
%
However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectowrise, LaraDB, GMAP, or SciDB building physical layouts with other constructs based in array programming~\cite{boncz2012vectorwise,hutchison2017laradb,  tsatalos1996gmap, stonebraker2013scidb}.
%
However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
%
Lastly, we mention SPIRAL, as a system that can express a structure and computation that none of the systems mentioned above can: a recursive FFT. 
%
SPIRAL focuses on recursively defined linear algebra and computations on the same (e.g. FFT matrices as tensor products of smaller matrices)~\cite{franchetti2018spiral,franchetti2009operator}.
%

>>>>

\paragraph{ Compilers for Structured Data:}
Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
%
The TACO compiler originally supported just plain einsum computations, but has been extended several times to support (single dimensional) local tensors, breaks via semi-rings, windowing, tiling, and convolution~\cite{kjolstad_tensor_2017, kjolstad_tensor_2019, senanayake2020sparse, henry_compilation_2021,won2023unified}.
%
Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures and tree like structures~\cite{kjolstad_tensor_2017, chou2018format, chou2022compilation}.
%
The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}.
%
Similarly, SDQL originally supported only hash tables, but has been extended with one format language that allows one to specify tensors as queries on several base storage types~\cite{schleich2023optimizing} and via another system that optimizing programs by taking advantage of predicates describing symmetries and other structures on a tensor~\cite{ghorbani2023compiling}.
%
Taichi is another language with a separate program and data structure specification, but Taichi focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
%
Outside of these works, the window of comparison fogs up.
%
The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
%
Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
%
However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectowrise, LaraDB, GMAP, or SciDB building physical layouts with other constructs based in array programming~\cite{boncz2012vectorwise,hutchison2017laradb,  tsatalos1996gmap, stonebraker2013scidb}.
%
However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
%
Lastly, we mention SPIRAL, as a system that can express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT. 
%
SPIRAL focuses on recursively defined linear algebra and computations on the same (e.g. FFT matrices as tensor products of smaller FFT matrices)~\cite{franchetti2018spiral,franchetti2009operator}.
%
%RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these
